\textls[-0]{\thetitle}

### 2023-03-19 canonicitymodel.tex
Reply and add new comments.

Now we prove canonicity (and consistency) for \TT using a logical-relations model.
We follow the reducibility argument of
\citet{kaposi2019gluing}, \citet{coquand2018canonicity}, and \citet{sterling2019algebraic}
to construct our model.
The meta\-languages of these prior models are based on QIITs, categories with
families \cite{dybjer1995internal}, and the generalized algebraic theory
\cite{cartmell1986generalised}, respectively.
Without exposing the reader to too many technical details, our meta\-language
should be understood an instance of any of the above logical frameworks.
%Our model works directly on the earlier introduced syntax for readability\footnote{We still
%take quotient, so ultimately our syntax should be considered as an instance of
%any of the above logical framework, without exposing the reader too many
%technical details}.

@@

\noindent
Canonicity is a basic criterion that should be met by a dependent type theory serving as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
We can even argue that, if this theorem is proven in a computable meta\-logic, 
then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
interpreter for closed terms of the boolean type.%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

@@

\begin{assumption}[Set-theoretic Universe Assumption] As a simplifying move, we assume an infinite hierarchy of Grothendieck universes $\Set{i}$ for $i \in \mathbb{N}$ in our ambient meta\-logic.
\end{assumption}

We can roughly consider each Grothendieck universe $\Set{i}$ as the $\Set{i}$ in Agda. 
\begin{itemize}
  \item Each $\Set{i}$ is closed under dependent function types and dependent
  pair types. For example, later, for our interpretation of dependent function
  types, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have
  $\denotesC{\TyPi{A}{B}} \in \Set{i}$.
  \item The universe hierarchy is cumulative, as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$.
  \item Thus if $\denotesC{A} \in \Set{i}, \denotesC{B} \in \Set{j}$, we will
  have $\denotesC{\TyPi{A}{B}} \in \Set{i} \cup \Set{j} = \Set{i \lcup j}$.
\end{itemize}

@@

%The difficult part of the proof-relevant logical-relations model is the
%treatment of propositions---
A proposition is encoded as a subsingleton data. For
example, $\{\star^1 \ | \text{ when } t \equiv \true  \} \dot{\cup} \{\star^2 \
| \text{ when } t \equiv \false  \}$ represents a disjoint union of two
subsingleton sets, where the first one is either a singleton if we have a proof of
$t \equiv \true$ or an empty set otherwise.
In other words,
$\star^1$ witnesses a proof that $t \equiv true$.
Here, $\star$, $\star^1$, $\star^2$ are just some arbitrary fixed elements.

Given the above model for MLTT, there should be a function ${\Pi}^c$ such
that ${\Pi}^c({\denotesC{A}},{\denotesC{B}}) = \denotesC{\TyPi{A}{B}}$. We will
use this function when defining the logical-relations model for the rest of \TT,
Type-theoretically speaking, this ${\Pi}^c$ uses the \emph{internal dependent
function type} of
the above model.
However, such a function $\Pi^c$ is not yet possible because the definition
$\denotesC{\TyPi{A}{B}}$ is not based solely on $\denotesC{A}$ and
$\denotesC{B}$,
but also on the syntax $\goodType{\Gm}{A}{}$ and $\goodType{\Gm, A}{B}{}$.

Thus, we define a new denotation $\denotesCC{S} \coloneq (S, \denotesC{S})$
that returns the syntax piece $S$ as well.\footnote{This is also called
glued interpretation in \citet{sterling2019algebraic}.}
Thus, we can have a
function $\Glued{\Pi}$ such that $\GluedPi{\denotesCC{A}}{\denotesCC{B}} =
\denotesCC{\TyPi{A}{B}}$ now that the syntax is available.
Similarly, there are functions $\Glued{\Sigma}$, $\Gpair{a}{b}$  and
$\GSubstExt{\gm}{t}$ for dependent pair types, dependent pairs, and
substitution extension (and more for other constructions).
Furthermore, given $\Glued{S}$ (i.e., the syntax and its semantic interpretation),
we use $\Gmodel{\Glued{S}}$ to mean the latter of the two.\YZ{why not just $\denotesC{S}$?}
% These will be useful in our \TT model because as syntactic translation hinted, \TT reuses a lot from MLTT. What's more, during the construction of the model for \TT, we will use $\Glued{(\cdot)}$ when we can.

>>>>

Now we prove canonicity (and consistency) for \TT using a logical-relations model.
We follow the reducibility argument of
\citet{kaposi2019gluing}, \citet{coquand2018canonicity}, and \citet{sterling2019algebraic}
to construct our model.
The meta\-languages of these prior models are based on QIITs, categories with
families \cite{dybjer1995internal}, and the generalized algebraic theory
\cite{cartmell1986generalised}, respectively.
Without exposing the reader to too many technical details, our meta\-language
should be understood an instance of any of the above logical frameworks---the difference is our formulation did quotient manually but if we use their framework, quotient is done automatically.
%Our model works directly on the earlier introduced syntax for readability\footnote{We still
%take quotient, so ultimately our syntax should be considered as an instance of
%any of the above logical framework, without exposing the reader too many
%technical details}.

@@

\noindent
Canonicity is a basic criterion that should be met by a dependent type theory serving as
a programming language or as a computational foundation for mathematics.
%\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
% We can even argue that, if this theorem is proven in a computable meta\-logic, 
% then by the Curry–Howard correspondence, the canonicity theorem provides a big-step
% interpreter for closed terms of the boolean type.
%\YZ{Can we say something about
%strong/weak normalization?}\EDJreply{
%  No. \\
%  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
%  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
%      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
%    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
%  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
%  4. Of course we can conjecture we have that normalization function \\
%  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
%  but we don't have termination (a runnable interpreter) for open terms.  \\ 
%  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
%}

@@

\begin{assumption}[Set-theoretic Universe Assumption] We assume an infinite hierarchy of Grothendieck universes $\Set{i}$ for $i \in \mathbb{N}$ in our ambient meta\-logic.
\end{assumption}

We can roughly consider each Grothendieck universe $\Set{i}$ as the $\Set{i}$ in Agda: 
\begin{itemize}
  \item Each $\Set{i}$ is closed under dependent function types and dependent
  pair types. For example, later, for our interpretation of dependent function
  types, when we have $\denotesC{A}, \denotesC{B} \in \Set{i}$, we will have
  $\denotesC{\TyPi{A}{B}} \in \Set{i}$.
  \item The universe hierarchy is cumulative, as $\Set{i} \in \Set{i+1}$ and $\Set{i} \subseteq \Set{i+1}$.
  \item Thus if $\denotesC{A} \in \Set{i}, \denotesC{B} \in \Set{j}$, we will
  have $\denotesC{\TyPi{A}{B}} \in \Set{i} \cup \Set{j} = \Set{i \lcup j}$.
\end{itemize}

@@

%The difficult part of the proof-relevant logical-relations model is the
%treatment of propositions---
Proof-relevant model treats proposition in a special way : \EDJ{Without this first sentence, this paragraph seems unmotivated.}
a proposition is encoded as a subsingleton data. For
example, $\{\star^1 \ | \text{ when } t \equiv \true  \} \dot{\cup} \{\star^2 \
| \text{ when } t \equiv \false  \}$ represents a disjoint union of two
subsingleton sets, where the first one is either a singleton if we have a proof of
$t \equiv \true $ or an empty set otherwise.
In other words,
$\star^1$ witnesses a proof that $t \equiv \true$.
Here, $\star$, $\star^1$, $\star^2$ are just some arbitrary fixed elements.

Given the above model for MLTT, there should be a function ${\Pi}^c$ such
that ${\Pi}^c({\denotesC{A}},{\denotesC{B}}) = \denotesC{\TyPi{A}{B}}$. 
Type-theoretically speaking, this ${\Pi}^c$ uses the \emph{internal dependent
function type} of
the above model. We hope to
use this function when defining the logical-relations model for the rest of \TT.
However, such a function $\Pi^c$ is not yet possible because the definition
$\denotesC{\TyPi{A}{B}}$ is not based solely on $\denotesC{A}$ and
$\denotesC{B}$,
but also on the syntax $\goodType{\Gm}{A}{}$ and $\goodType{\Gm, A}{B}{}$.

Thus, we define a new denotation $\denotesCC{S} \coloneq (S, \denotesC{S})$
that returns the syntax piece $S$ as well.\footnote{This is also called
glued interpretation in \citet{sterling2019algebraic}.}
Thus, we can have a
function $\Glued{\Pi}$ such that $\GluedPi{\denotesCC{A}}{\denotesCC{B}} =
\denotesCC{\TyPi{A}{B}}$ now that the syntax is available.
Similarly, there are functions $\Glued{\Sigma}$, $\Gpair{a}{b}$  and
$\GSubstExt{\gm}{t}$ for dependent pair types, dependent pairs, and
substitution extension (and more for other constructions).
Furthermore, given $\Glued{S}$ (i.e., the syntax and its semantic interpretation),
we use $\Gmodel{\Glued{S}}$ to mean the latter of the two.\YZ{why not just $\denotesC{S}$?}\EDJreply{Using superscript C is shorter and more precise -- we are not "reconstructing the C model from the syntax, but taken out the C model directly from the pair". Look at the canonicity model where superscript c is used, if change to $\denotesC{-}$ on them, it is not really well-founded recursion, technically speaking (only equivalent to one).}
% These will be useful in our \TT model because as syntactic translation hinted, \TT reuses a lot from MLTT. What's more, during the construction of the model for \TT, we will use $\Glued{(\cdot)}$ when we can.

### 2023-03-19 metatheory2.tex
Took a pass over appendix. A question was added.
edit
Reply and add new comments.

\Citeauthor{debruijn1964} indices and explicit substitutions make details about
binders and substitutions explicit.
%The change of the syntax formulation is important to the simplicity of our
%proof, as we want to handle the proof about substitutions and terms
%uniformly---thus making substitution as part of the syntax (explicit
%substitution) will be helpful.

The form \var{n} represents a variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

Universe levels address the size issue---it is unsound to have a set of all sets
(or a universe of all types).
The level of a universe specifies ``how large'' that universe is.
%that is basically what universe level is measuring.

@@

\subsection{Proof Relevant Logical Relation for Canonicity}
\input{canonicitymodel.tex}

>>>>

\Citeauthor{debruijn1964} indices and explicit substitutions make details about
binders and substitutions explicit.
%The change of the syntax formulation is important to the simplicity of our
%proof, as we want to handle the proof about substitutions and terms
%uniformly---thus making substitution as part of the syntax (explicit
%substitution) will be helpful. 
\EDJ{Why comment out this sentence?}
The form \var{n} represents a variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

Universe levels address the size issue---it is unsound to have a set of all sets
(or a universe of all types)~\cite{hurkens1995simplification}\EDJ{Why remove this citation?}.
The level of a universe specifies ``how large'' that universe is.
%that is basically what universe level is measuring.

@@

\subsection{A Proof Relevant Logical-Relations Model for Canonicity}
\label{sec:canonicity-model}
\input{canonicitymodel.tex}

### 2023-03-19 syn-translate.tex
Reply and add new comments.

\begin{align*}
  {\Sigr{\Gamma}{(n+1)}{j}} &= 
    \Sigma~(\Sigr{\Gamma}{n}{j})
          ~(\Sigma~\cU_j \\
          & \quad \quad ~(\Sigma~(\TyPi{{\mathscr{P}({\sub{\var{0}}{\SubstWeak{1}}})}}{\El{\sub{\var{0}}{\SubstWeak{1}}}}) \\
          & \quad \quad \quad \TyPi{\El{\sub{\var{0}}{\SubstWeak{1}}}}{\cU_j})) \\
  \Sigr{\Gamma}{0}{j} &= \top \\
  \mathscr{P}(\lsig') &=  \top \quad \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{0}{j}}  \\
  \mathscr{P}(\lsig') &=  
    \Sigma~(\fst{\lsig'})~(\El{\app{"snd"^3~\lsig'}[(\SubstWeak{1}, \app{\fst{("snd"^2~{\lsig'})}})]}) \\ & \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n+1}{j}} \\
\end{align*}

@@

We omit validating the equational rules ($\beta$, $\eta$, and substitution) here.
However, when we mutually recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}{}} $, and
$\denotesT{\cP}$ above, we actually have to prove the substitution laws
$\sub{(\TyTkg{\lsig})}{\gm}\equiv \TyTkg{\sub{\lsig}{\gm}}$ and $\sub{(\Sigr{\Delta}{n}{})}{\gm}
\equiv \Sigr{\Gamma}{n}{}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.

>>>>

\begin{align*}
  {\Sigr{\Gamma}{(n+1)}{j}} &= 
    \Sigma~(\Sigr{\Gamma}{n}{j})
          ~(\Sigma~\cU_j \\
          & \quad \quad ~(\Sigma~(\TyPi{{\mathscr{P}({\sub{\var{0}}{\SubstWeak{1}}})}}{\El{\sub{\var{0}}{\SubstWeak{1}}}}) \\
          & \quad \quad \quad \TyPi{\El{\sub{\var{0}}{\SubstWeak{1}}}}{\cU_j})) \\
  \Sigr{\Gamma}{0}{j} &= \top \\
  \mathscr{P}(\lsig') &=  
    \Sigma~(\fst{\lsig'})~(\El{\app{"snd"^3~\lsig'}[(\SubstWeak{1}, \app{\fst{("snd"^2~{\lsig'})}})]}) \\ & \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{n+1}{j}} \\
  \mathscr{P}(\lsig') &=  \top \quad \text{ when } \goodTerm{\Gm}{\lsig'}{\Sigr{\Gm}{0}{j}}  \\
\end{align*}

@@

We omit validating the equational rules ($\beta$, $\eta$, and substitution) here.
Note that, when we mutually recursively define the type
${\Sigr{\Gamma}{n}{}}$ (of signatures $\denotesT{\goodSig{\Gamma}{t}{n}}$), and
$\denotesT{\cP} = \mathscr{P}$ above, we actually have to prove the two substitution laws
$\sub{(\TyTkg{\lsig})}{\gm}\equiv \TyTkg{\sub{\lsig}{\gm}}$ and $\sub{(\Sigr{\Delta}{n}{})}{\gm}
\equiv \Sigr{\Gamma}{n}{}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.
